---
title: "Setup Check"
author: "YOUR NAME"
date: today
format: html
---

## Instructions

1. Replace `"YOUR NAME"` above with your actual name
2. Render this document (Ctrl+Shift+K or Cmd+Shift+K)
3. If successful, commit and push this file (the rendered output is ignored by Git)

## Environment Check

The code below verifies that your R packages are installed correctly.

```{r}
#| message: false

# Load required packages
library(tidyverse)
library(gtsummary)
library(gt)
library(here)
library(NHANES)
```

If the above runs without errors, your packages are correctly installed.

## Quick Test

```{r}
#| tbl-cap: "Baseline characteristics of NHANES participants"

NHANES |>
  select(Age, Gender, BMI, BPSysAve) |>
  head(100) |>
  tbl_summary(
    by = Gender,
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    missing = "no"
  )
```

## Session Info

```{r}
sessionInfo()
```

## Confirmation

After rendering this document:

- [ ] The HTML file was created without errors
- [ ] The table above displays correctly
- [ ] I have committed and pushed my changes to the `.qmd` file

**You're ready for the workshop!**
